---
title: "rHFA Guide"
author:
   - name: "Patrick Ewing"
     role: ["aut", "cre"]
     date: "2024-10-11"
   - name: "Eric Killian"
     role: ["aut"]
     date: "2022-09-12"
   - name: "Michael Kantar"
     role: ["aut"]
     date: "2024-10-18"
   - name: "Tom Glancy"
     role: ["edt"]
     date: "2024-10-29"
output: 
  html_document:
    toc: true
    toc_float:
      collapsed: false
editor_options: 
  markdown: 
    wrap: 72
vignette: >
  %\VignetteIndexEntry{rHFA Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

## Home Field Advantage in Plant Breeding

Home Field Advantage (HFA) is the fitness gain an entry (genotype)
realizes by growing in its home location, which is empirically defined
as the location of highest fitness relative to other entries ([Ewing *et
al*.,
2019](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0227079)).
At the population level, i.e. averaged across members of a population,
it can indicate population adaptation strategy and, for bred
populations, the preferences of the breeder. We also can evaluate how
this adaptation has changed across time among populations. Finally, we
can use it to identify specialist and generalist genotypes. When
combined with, for example, kinship, heritability ([MacQueen *et al*.,
2021](https://acsess.onlinelibrary.wiley.com/doi/full/10.1002/csc2.20694]))
or genetic potential ([Ewing *et al*.,
2024](https://acsess.onlinelibrary.wiley.com/doi/full/10.1002/csc2.21168)),
HFA allows novel insights into population structure and breeding
strategy.

### How Home Field Advantage is Calculated

Overall home field advantage is conceptualized as the average additional
yield benefit conferred to a variety when growing at it's home location
after accounting for site, year, site-by-year, and genetic potential.
Begin with:

Eq. 1:
$phenotype_{ijk} = genotype_i + site_j + year_k + site_j \times year_k + is\_home_{ij} + \epsilon_{ijk}$

Where *is_home* is an indicator for whether site *j* is the home site
for variety *i*; its coefficient is the home field advantage (HFA)
across genotypes within the population. This metric quantifies the
benefits of environmental specialization using “home field advantage”
(HFA), which was defined as the fitness gain an organism has in its
native soil ([Blanquart *et al*.,
2013](https://doi.org/10.1111/ele.12150)). Modern crop populations
lack an obvious home location, so [Ewing *et al*.
(2019)](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0227079)
redefined HFA as the yield advantage that a crop variety receives at its
best performing, “home” location, where “best” is relative to other
varieties, regardless of the location of the breeding program. This
definition contrasts with a variety's origin—the target region for which
it was bred.

Since "home" was picked as a good site, we expect this to be a positive
coefficient. The real question is, was HFA larger than expected. This is
tested using permutation. In this case, values of *phenotype* are
permuted within site-year. By calculating HFA on this permuted data a
bunch of times, we can estimate what we'd expect HFA to be based on
chance and the variation in the dataset. Larger-than-expected HFA are
indicative of local adaptation, while smaller-than-expected HFA are
indicative of broad adaptation.

Finally, with wild data - where sites and genotypes may change or may
not be fully crossed - finding a "best" site is more challenging. rHFA
provides a few ways to identify the home site:

-   A best linear unbiased predictor (BLUP) of relative performance in a
    given site-year. This is via a random intercept model using
    `nlme::lme()`. Because it accounts for certainty (i.e. more measures
    of each hybrid-site combination) it's best for unbalanced, sparse,
    and highly variable data. But it's also slow.

<!-- -->

-   Raw means (within site across year). Fast, for balanced,
    well-behaved (i.e. normal-ish) data with a deeper time series at
    each site and for each genotype.
-   Median. Fast, for highly and non-normally variable, relatively dense
    data with a deeper time series at each site and for each genotype.
-   You can also provide your own, generic function. See
    `help('id_home')`.

Note that home site assignments and effect sizes will differ among the
modeling approaches, especially for sparser datasets.

# Guide

## Step 1: Acquire data

Data can be acquired from a number of common garden-like designs. They
must be in long format containing the plant variety and data from
multiple years at each location.

```{r}
library(rHFA)

in_corn <- system.file('extdata', 'illinois_corn_yield.csv', package = "rHFA")
corn <- read.csv(in_corn)

# convert columns to factors
facts <- c('YEAR', 'COMPANY', 'HYBRID', 'SITE')
corn[facts] <- lapply(corn[facts], as.factor)

head(corn, 10)
corn_full <- corn  # to access the original data later
```

```{r}
str(corn)
```

This dataset is a subset of the corn variety trials conducted by the
University of Illinois Extension. Yield is in bushels per acre. For
those keeping track, one standard bushel is 56 pounds.

Note that this is a ***tall*** dataframe. Columns and
`corresponding arguments` in `rHFA` functions are:

-   YEAR (*factor*): `year` instance of data collection - likely year.
-   COMPANY (*factor*): `popn` differentiates subpopulations within the
    dataset. In this case, company that entered the hybrid.
-   HYBRID (*factor*): `geno` Hybrid (genotype) identifier.
-   SITE (*factor*): `site`, the location of data collection.
-   YIELD (*numeric*):`pheno`, the phenotype of interest. Here, a
    hybrid's average yield across replicates within the site-year.

The original design is somewhere between an augmented block and a common
garden experiment: a population of commercial hybrids was grown across
locations, only the "checks" were grown in each location in a given
year, but checks, hybrids, and even locations changed across the
experiment. We have already subsetted the dataset to include only those
checks.

The rate of turnover varies among hybrids and sites. As a result, this
is a very sparse, unbalanced dataset: sites and hybrids differ across
years. For example, here is a count of the number of years each site was
used.

```{r}
count_instances(corn, 'SITE', 'YEAR')
```

You can see sites were used in as few as one year, while some were used
in all 19 years.

```{r}
count_instances(corn, 'HYBRID', 'YEAR') |>
  summary()
```

Most hybrids were used in just one year and none were grown in all
years.

This type of challenge is extremely common with public data and causes
problems in reliable identification of home site - which is the critical
step of HFA analysis. The most robust home site assignments will be
robust to weather variation, so multiple years is essential. rHFA
provides access to different modeling approaches to identify "home".
Still, you probably will want to filter your data a bit to ensure
adequate replication. rHFA provides tools to filter the dataset to a
denser subset.

### Notes about checking your data

It is important for you to make sure that you examine your data for
missing values, for consistency of names across sites, and for
consistency of varieties across sites. Doing checks for outliers and
potentially incorrect values is also important. Do this **before**
filtering to a viable subset!

## Step 2: Filter data to a viable subset

Both sites and hybrids need to be present for at least two years for the
HFA analysis to work, although more is better. Representation at sites
needs to be complete, as well. This is because home site assignment is
in part dependent on the other lines grown.

We will:

1.  Select locations that were used at least 2 times
2.  Select checks that were grown in at least 2 years
3.  Select location-years with at least 2 lines

Of course, 2 is the minimum for each of these. If you run into errors,
especially in the `permute_hfa()`, try increasing `min_times`.

We can filter in two ways using `rHFA` functions:

### Option 1: Manual filter

First, we can manually for each criterion with `filter_instances()`.
This allows filtering to different counts for each combination of
hybrid, site, and year. For example, to filter hybrids to those grown in
at least two years leaves us with:

```{r}
corn <- filter_instances(
  data=corn,
  group_col='HYBRID',
  rep_col='YEAR',
  min_times=2,        # default is 2
  na.rm=TRUE          # ignore missing data instances
)

count_instances(corn, 'HYBRID', 'YEAR') |>
  summary()
```

This gives us `r length(levels(corn$HYBRID))` hybrids to work with,
versus `r length(unique(corn_full$HYBRID))` in the original data.

Doing the same for site:

```{r}
corn <- filter_instances(
  data=corn,
  group_col='SITE',
  rep_col='YEAR',
  min_times=2,
  na.rm=TRUE
)

count_instances(corn, 'SITE', 'YEAR')
```

Leaving `r length(unique(corn$SITE))` sites, down from
`r length(unique(corn_full$SITE))`.

And we need at least 2 hybrids grown in each site-year.

```{r}
corn$SITEYEAR <- with(corn, 
                     paste(SITE, YEAR, sep='_'))

corn <- filter_instances(corn,
                        group_col='SITEYEAR',
                        rep_col='HYBRID',
                        min_times=2,
                        na.rm=TRUE)
```

We now have a dataset with `r nrow(corn)` observations, versus
`r nrow(corn_full)` of the raw dataset.

Consider repeating these filtering steps a couple of times until the
final number of observations is stable. See the next section.

### Option 2: Autofilter

rHFA also includes an `autofilter_instances()` function that filters
based on these three criteria (site-year, hybrid-year, and hybrids
within site-year). To demonstrate, we'll increase the minimum number of
instances to 3 using the `min_times` argument.

```{r}
test <- autofilter_instances(corn, 
                             site='SITE', 
                             year='YEAR', 
                             geno='HYBRID', 
                             min_times=3, 
                             max_cycles=999)
```

Note that `autofilter_instances()` repeated itself for a total of two
cycles. This is because filtering, for example, hybrids within site-year
may reduce the occurrence of site-years to fewer than `min_times`.
`autofilter_instances()` will repeat itself until the dataset is stable.

The data is now ready for home field advantage analysis. You can skip
right to Step 5, `permute_HFA()`, to estimate and test HFA immediately.
The following builds into that.

## Step 3: Permute HFA

We're ready to run the HFA analysis. You really just need one function
to do this: `permute_hfa()`. The default is to run this at the
population level. For meta-populations with sub-populations (for
example, domestication events of dry bean or individual breeding
programs), specify a column that differentiates them.

```{r}
pophfa <- permute_hfa(corn,
                     level='population',  # default
                     site='SITE',
                     year='YEAR',
                     geno='HYBRID',
                     pheno='YIELD',
                     population=NA, # specify to run separately for each sub-population.
                     times=9, 
                     method='blup',
                     parallel=TRUE, 
                     seed=73956)
```

For speed reasons, we'll limit ourselves to 9 permutations (plus the
observed data). In practice, use *at least* 99, preferably 999. If you use
`method="blup"`, you may hit some errors. Try increasing `min_times` in
`autofilter_instances()` or using a different `method`.

Now, look at results. `permute_hfa()` returns a list with three objects:

1.  `.$home_field`, which contains HFA effects and the results of a
    permutation test.
2.  `.$perms`, which contains the permutations
3.  `.$formula`, the formula used to calculate HFA.

```{r}
pophfa$home_field
```

For the entire population, the observed HFA was 12.1 bushels per acre, a
`r round(pophfa$home_field[1, 1]/mean(corn$YIELD), 3)*100`% increase
over the average corn yield in this dataset. However, this was about 1.3
bushels per acre less than expected (90% confidence interval: -2.2 to
-0.2 bushels per acre). Every permutation's HFA was larger than the
observed HFA (*p* = 0), which you can verify by looking at the
permutations themselves:

```{r}
pophfa$perms
```

You can also access the formula used to calculate HFA:

```{r}
pophfa$formula
```

And finally, the data including a home site assignment:

```{r}
head(pophfa$data)
```

Three new columns are here for your inspection:

1.  rel_YIELD, which is the relative yield of a hybrid within a
    site-year.
2.  blup_rel_YIELD, the expected relative yield of that hybrid at that
    site-year as determined by `method="blup"`. 3. is_home, which flags
    sites as home or not.

## Step 4: Variance explained

You might wonder how important this HFA is relative to other factors,
say, location or genotype. The `get_ss()` function will do this.

```{r message=FALSE}
mod <- lm(pophfa$formula, 
          data=pophfa$data)
ss <- get_ss(mod)
ss
```

Home site explains

`r round(ss['is_home', 'Percent Var'], 3)`% of total variance. This may not seem
like much, but it is `r round(ss['is_home', 'Sum Sq']/sum(ss[c('is_home', 'Residuals'), 'Sum Sq']), 3)*100`% of otherwise-unexplained GxE (residuals + is_home). 

P-values here are from the package, `car`, based on the indicated *F*
test. `get_ss()` is just a wrapper for `car::Anova()`.

## Step 5: ID home site

To quickly match home site with a genotype, use `get_home_site()`:

```{r}
get_home_site(pophfa$data, 
              geno='HYBRID', 
              site='SITE') |> 
  head()
```

SITE is the home site of the indicated hybrid, while home_years is the
number of years that hybrid was grown at that site.

# Additional HFA analyses

## Identifying home without permuting

If you just want home sites and don't care about significance testing,
use `id_home()`, which produces the `data.frame` of the permutations
results (`pophfa$data`).

```{r message=FALSE, warning=FALSE}
corn_home <- id_home(
  data=corn,
  site='SITE',
  year='YEAR', 
  geno='HYBRID', 
  pheno='YIELD',
  method='blup',
  verbose=FALSE
) 
subset(corn_home, HYBRID == corn[110, 'HYBRID']) |>
  head(10)
```

As a gut check, let's look at relative yield at home sites.

```{r}
aggregate(rel_YIELD ~ SITE + HYBRID, 
          data=corn_home,
          subset=is_home,
          FUN=mean) |>
  summary()
```

Lines at home tend to perform better than average at that location (V1 =
rel_YIELD \> 0).

## Specialists and generalists

We can also compute a genotype-specific HFA, which we can then use to
identify specialists (higher HFA than expected) and generalists (lower
HFA than expected).

```{r}
genohfa <- permute_hfa(corn,
                       level='genotype',
                       site='SITE',
                       year='YEAR',
                       geno='HYBRID',
                       pheno='YIELD',
                       times=9, 
                       method='blup',
                       parallel=TRUE, 
                       seed=73956)
```

The default test values here are whether the observed HFA is different
from permutations:

```{r}
genohfa_test <- genohfa$home_field
head(genohfa_test)
```

Based on raw p-values, we find these genotypes have an HFA different
from zero:

```{r}
tt <- genohfa_test$p_val < 0.05
genohfa_test[tt, ]
```

Since all HFA are smaller than expected, we these are probably
*generalists*, but we might also want to adjust these p-values as we
have a lot of comparisons. In practice, do this with more permutations.

```{r}
genohfa_test$p_adj <- p.adjust(genohfa_test$p_val, 
                               method='fdr')

tt <- genohfa_test$p_val < 0.05
genohfa_test[tt, ]
```

HFA is less than expected, so we infer (but have not tested) that these
are *generalists*. We can test this using a one-tailed test for each
generalists and specialists.

```{r}
geno_hfa <- specialist_test(genohfa) |>
  generalist_test()
head(geno_hfa$home_field)
```

p_specialist and p_generalist are the *p* values, i.e. that the null
hypothesis (specialist/generalist due to chance) is true.

## Spatial variation in HFA
To identify the average HFA conferred at each site (for home genotypes), set 
`level="site"`. 

```{r}
sitehfa <- permute_hfa(corn,
                       level='site',
                       site='SITE',
                       year='YEAR',
                       geno='HYBRID',
                       pheno='YIELD',
                       times=9, 
                       method='blup',
                       parallel=TRUE, 
                       seed=73956)
```

```{r}
sitehfa |>
  specialist_test() |>
  generalist_test() |>
  _$home_field
```

These are the advantages conferred to home lines at each site. Again,
results suggest generalist populations. Running `specialist_test()` and `
`generalist_test()` on this provide some nuance:

-   Mt. Morris may be an environmentally distinct location, as HFA is
    higher than expected.
-   Goodfield may be especially representative, with an extremely low
    HFA.

## Temporal variation in HFA

```{r}
yearhfa <- permute_hfa(corn,
                       level='year',
                       site='SITE',
                       year='YEAR',
                       geno='HYBRID',
                       pheno='YIELD',
                       times=9, 
                       method='blup',
                       parallel=TRUE, 
                       seed=73956)
```

```{r}
yearhfa$home_field
```

We don't see an obvious trend here, but plotting may help:

```{r}
yhfa <- yearhfa$home_field
plot(as.numeric(yhfa$YEAR), 
     yhfa$observed,
     xlab='Year',
     ylab='HFA')
points(as.numeric(yhfa$YEAR),
       yhfa$observed - yhfa$diff_q50, col='red')
legend('topright',
       c('expected', 'observed'),
       col=c('black', 'red'),
       pch=1)
```

No obvious trend - if anything, it may be slighlty down. Of course we
can test this:

```{r}
lm(observed ~ as.numeric(YEAR), 
   data=yearhfa$home_field) |>
   summary()
```

HFA has not changed across years.

------------------------------------------------------------------------

# *The following sections are under construction...*

------------------------------------------------------------------------

# Extensions {style="color: lightgray"}

## Distances among home sites {style="color: lightgray"}

Two functions measure the distance between home sites of genotype pairs.
`home_distance_spatial()` measures the spatial distance using lat-long
coordinates. `home_distance_environmental()` will calculate any distance
based on any columns, for example, weather data.

If you get errors, check that:

1.  The site attributes dataframe should have the same `site` column
    name as the phenotypic dataframe.
2.  All sites in the phenotype dataframe should be present in the site
    attributes dataframe.

```{r}

# corn_sites = system.file('extdata', 'illinois_sites.csv', package = "rHFA")
# corn_sites = read.csv(corn_sites)
# 
# names(corn_sites) = gsub('City', 'SITE', names(corn_sites))
# corn_sites[, 'SITE'] = tolower(corn_sites[, 'SITE'])

# home_dist = home_distance_spatial(
#   data=corn,
#   locations=corn_sites,
#   geno='HYBRID',
#   site='SITE',
#   lat='Lat',
#   long='Lon'
# )
# 
# home_dist[1:5, 1:5]
```

Note the distances are in kilometers.

The generic

```{r}
# envi_dist = home_distance_environmental(
#   data=corn,
#   locations=corn_sites,
#   geno='HYBRID',
#   site='SITE',
#   vars=c('bio_01', 'bio_17', 'BLDFIE_M_sl1_1km_ll', 'SLTPPT_M_sl1_1km_ll'),
#   scale=TRUE,
#   method='euclidean'
# )
# 
# envi_dist[1:5, 1:5]
```

These units are fairly meaningless.

## Kinship among genotypes and other covariates {.color: .lightgray}

Add a "covariates" argument to allow inserting a generic covariate to
the equations. ex. covariate="is_home\*kinpc1"

```{r}
#
#
#

```

------------------------------------------------------------------------

# Graphics {style="color: lightgray"}

Arrows plot.

HFA vs yield potential.
